State of the Practice for Lattice Boltzmann Method Software

### 2022-01-04 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Complete pass through LBM SOP - start of reorg of current, best and future practices

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

@@

Seventeen of the software packages include a requirements specification artifact
or explicitly reference domain theory, often only the latter. Software packages
that distribute requirements specification information, such as DL\_MESO (LBE),
generally keep it brief and include it within other documentation. This artifact
is often found within a user manual, on a web-page, or is mentioned in related
publications. In the latter case the user may need to spend significant time to
find this information.

@@

@@

Improving the software engineering education or experience of developers is also
an idea that was brought up by several developers.

@@

This section answers the research questions listed in Section~\ref{secIntro}
using the quantitative data presented in Section~\ref{AHPresults}, qualitative
data presented in Section~\ref{interviewresults}, additional data from software
package repositories and artifacts, and domain expert feedback. Each subsection
answers one of the research questions. Software package artifacts are examined
in Section~\ref{artifacts}, tools are listed in Section~\ref{tools}, development
principles, processes, and methodologies are discussed in Section
~\ref{prinprocmeth}, development pain points are explored in Section
~\ref{painpoints}, and our recommendations for improving software qualities are
presented in Section~\ref{qualityrecommentations}. A comparison between our
designations of the software packages to community rankings is made in Section
~\ref{comparison}. Finally, threats to the validity of this assessment are noted
in Section~\ref{threats}.

@@

\begin{table}
\begin{center}
\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
\toprule
\textbf{Common} & \textbf{Less Common} & \textbf{Rare} \\
\midrule
Authors / Developers List & Change Log / Release Notes & API Documentation\\
Bug Tracker & Design Documentation & Developer / Contributor Manual\\
Dependency Notes & Functional Spec. / Notes & FAQ / Forum\\
Installation Guide / Instructions & Performance Information & Verification and Validation Plan\\
Requirements Spec. / Theory Notes & Test Plan / Script / Cases & Video Guide (including YouTube)\\
List of Related Publications & User Manual/Guide & \\
Makefile / Build File & Version Control & \\
README File & & \\
License & & \\
Tutorial & & \\
\bottomrule
\end{tabular}
\caption{Artifacts Present} \label{artifactspresent}
\end{center}
\end{table}

@@

The top four AHP ranked packages have most of the less common artifacts. At the
time of data collection, only one (Palabos) of the four packages did not have a
user manual or guide, but there was a broken link on the package website
indicating that such an artifact might exist. This broken link was later fixed,
but this is not reflected in our data because it was not present at the time of
data collection. Despite the broken link, Palabos does have a detailed and
informative website. Another one (Ludwig) of the top four packages does not
appear to have publicly visible design documentation. A third package (OpenLB)
from the list does not appear to use a version control system. It is possible
that such a system is used by the package, since its website notes package
version numbers, but the artifacts do not explicitly state the use of such a
system.

@@

@@

A developer of pyLBM noted that their small development team has a lack of time
to implement new features. Small development teams are common for LBM software
packages. Team members are almost always part of the same institute or already
know each other from other projects. External contributions are rare despite
many of the projects accepting them. Aside from on-boarding new developers, time
constraints could be mitigated by increasing developer efficiency, which could
be addressed in several ways, including by improving the quality of
documentation, or incorporating automatic code generation.

@@

A lack of software development experience was noted by the developer of TCLB.
Many of the team members on their project are domain experts and there can be a
steep learning curve before these team members contribute good quality source
code. It was further noted that this has been somewhat addressed, as the code
has been re-written to best ensure ease of future contributions.

@@

The same developer also noted that there is a lack of incentive and funding in
academia for developing widely used scientific software. The importance of
funding for scientific software has been discussed in
\cite{gewaltig2012quality}, which notes that ``software tools are developed and
maintained only for as long as there is explicit or implicit funding''. The
developer further commented that there are no journals that publish such
scientific software source code. However, there are ways to get such source code
cited. Work has been done to address this in \citep{smith2016software}, which
presents a set of software citation principles and discusses ``how they could be
used to implement software citation in the scholarly community''
\citep{katz2019software}.

@@

Scientific software is often developed in-house by the very researchers that
temporarily use it in their own research. Empirical studies of such
``professional end-user development'' of SCS is noted in \citep{segal2007end}.
This kind of software has a defined user and purpose, and often does not meet
the standards that would be required by external users. It has been categorized
as a ``private tool'' by \citep{gewaltig2012quality}, which notes that despite
often being made freely available, ``it is not always clear that it is
sufficiently mature in terms of domain coverage, validity, documentation or
usability, to be useful to other researchers''. It is less common for such ad
hoc scientific software to become ``user-ready software'', which ``is not only
research-ready, but should have most of the attributes commonly expected of
commercial software products including broadness of scope, robustness,
demonstrable correctness and adequate documentation''
\citep{gewaltig2012quality}. As software becomes user-ready, it can become
commercialized and closed-source.

@@

Setting up parallelization was also noted as a technical pain point by one of
the developers, and the introduction of continuous integration by another.
Software development knowledge, and automatic code generation, could mitigate
such pain points. As already noted, many of the developers are domain experts
and not professional software developers. The developer of TCLB noted that
eliminating equivalent statements using macros had helped improve the quality of
their source code, specifically helping with reusing code to run on both the CPU
and GPU.

@@

Difficulties with ensuring correctness were also noted by several developers.
They indicated that tests are run on all new source code additions, testing both
individual modules and the system to verify correctness. These tests compare the
package output to known correct output using test cases. The developer of TCLB
commented that the amount of testing data that is needed for some cases is a
problem as free testing services do not offer capabilities to store and process
such large amounts of data, and in-house testing solutions needed to be created
to address this limitation. The solution for this has been to limit the size of
the testing problems, and to run tests in small batches with few iterations.

@@

A few obstacles related to users were found. Several developers noted that users
sometimes try to use incorrect LBM method combinations to solve their problems.
Furthermore, some users think that the packages will work out of the box to
solve their cases, while the packages require both a good understanding of CFD
and an understanding of the requirements for formulating problems in the
individual packages, which can be a significant endeavor. These software
packages are not like commercial software packages. They are generally set up to
solve specific research problems, and are often primarily used by their
developers. While they are modifiable to solve similar problems, these
modifications are not trivial. Better documentation, with attention on
traceability, and automatic code generation are suggested when designing
software for change, and would help with these modifications. So far this
problem of on-boarding new users has been addressed by updating the
documentation to better inform users of the underlying LBM theory and package
requirements. Similar issues with LBM parameters were noted by another
developer. Updating the user interface to better explain theoretical principles,
as well as test user input for compatibility, was the implemented solution. As
noted above, sometimes frequently asked questions on the underlying theory and
on how to use the software are answered in the documentation.

The interviewed developer of ESPResSo commented that parts of their package's
source code had been refactored to Python to help address usability issues.
Python was perceived as a much more usable language, and it would be easy for
future users and developers to learn and understand the source code.

@@

A few potential future obstacles were noted. The developer of ESPResSo noted
that their source code had been written with a specific application in mind and
that due to this there was too much coupling between components of the source
code. This results in technical debt, having an impact on future modifiability
and reusability when trying to extend the software, and the code would need to
be refactored.

As noted above, difficulties with ensuring future correctness could also arise.
As new methods and functionality is added into the software, new test cases and
test data will need to be developed.

@@

Interviewees commented that documentation is important and that its quality
could be improved. As already noted, there is often no time or funding for
maintaining quality documentation for software that is rarely used outside of
the development team. Furthermore, the documentation generally only provides a
shallow overview of the underlying CFD theory. Users would be well advised to
already be familiar with these topics, or they should spend significant time
referencing theory resources. The documentation instead generally focuses on
explaining how to use the software. It is of course not feasible for package
documentation to address the underlying physics topics in detail, so it is
advised that the package documentation links to resources that better explain
the underlying theory. Sometimes, frequently asked questions about the
underlying theory are answered in the documentation. OpenLB has an artifact for
such questions.

@@

\subsection{What is the current software quality for the projects?}
\label{qualityrecommentations}

This subsection answers the research question: For research software developers,
what specific actions are taken to address software qualities?

The following points regarding software quality should be considered when
developing LBM software packages. These points are based on developer
interviews, SCS literature, what was found to have worked for packages that were
designated as high quality in this assessment.  Our recommendations are not
lists of what should have been done in the past, or what should be done now;
they are just suggestions for consideration in the future.

@@

\begin{itemize}
	\item Include OS compatibility, including specific OS versions. This was
	done by several top ranked packages, including ESPResSo and LUMA.
	\item Provide complete installation instructions. This was done by all of
	the top five ranked packages.
	\item Installation instructions should be written as if the user does not
	have any dependencies installed and is installing on a clean OS. This was
	done by several top ranked packages, including ESPResSo, OpenLB, and
	Palabos. This can be tested on a clean environment using a virtual machine.
	\item List all dependencies in the installation instructions, dependency
	versions, and how to install them. This was done by several top ranked
	packages, including ESPResSo, LUMA, OpenLB, and Palabos.
	\item If possible, automate the installation of dependencies. Use tools such
	as Make. This was done by all of the top five ranked packages.
	\item Automate the installation process as much as possible. Use tools such
	as Make. This was done by all of the top five ranked packages.
	\item The installation instruction should only be in one location. This was
	done by several top ranked packages, including ESPResSo, OpenLB, and
	Palabos.
	\item Include descriptive error messages for errors encountered during
	installation. This was done by several top ranked packages, including Ludwig
	and LUMA
	\item Provide a way to validate the installation. This can be done using a
	custom script, or a test case that specifies expected output. This was done
	by several top ranked packages, including OpenLB, Ludwig, LUMA, and Palabos.
	\item Provide instructions for uninstalling the software. These were
	included with pyLBM.
\end{itemize}

@@

\begin{itemize}
	\item Use a requirements specification document. This is suggested in SCS
	literature, and several of the top ranked packages had such a document or
	reference to theory manuals. A potential template is presented in
	\citep{smith2005new}. 
	\item Make public (on package website or GitHub) the requirements
	specification document, or explicitly reference the domain theory that the
	software is designed from.  This was done by several top ranked packages,
	including ESPResSo, Ludwig, LUMA, and OpenLB.
	\item Ensure the above information is easy to find. Consider adding it to
	the user manual. The information is easy to find in most top ranked
	packages.
	\item Development teams should include both domain experts and experienced
	software developers. This suggestion is based on developer comments.
	\item Use and make public (on package website or GitHub) detailed
	documentation. Consider using automatic document generation tools like
	Doxygen, Drasil, or Sphinx. This was done by all of the top five ranked
	packages.
	\item Provide detailed tutorials that include expected output, like the
	\href{https://www.walberla.net/doxygen/index.html}{waLBerla tutorials}.
	\item Use unit tests during development and make them public (on package
	website or GitHub). This was available for Ludwig.
	\item Modularize the source code, separate components, hide information
	behind well defined interfaces. This is suggested in SCS literature, and in
	developer comments.
	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and
	processes during development. This was done by top ranked packages ESPResSo
	and Ludwig.
\end{itemize}

@@

\begin{itemize}
	\item Include descriptive error messages where appropriate. This was done by
	most packages that encountered a fault.
	\item In case automatic installation of dependencies fails, the system
	should indicate to the user what dependencies need to be installed manually.
	This was done by many packages, including ESPResSo++, Ludwig, LUMA, pyLBM,
	TCLB.
	\item The packages should include detailed tutorials, including
	dependencies, expected output, and any additional supplementary
	documentation that may be required. This was done by many packages,
	including waLBerla, Palabos, MechSys, LUMA, and pyLBM.
\end{itemize}

@@

\begin{itemize}
	\item Keep artifacts updated. This is suggested as good software maintenance
	practice.
	\item Ideally include most of the common and less common artifacts listed in
	Section~\ref{artifacts}.
	\item Include version numbers and release notes for all major source code
	and artifact releases. The top five ranked software packages include release
	notes.
	\item Have a defined process for accepting contributions, and make public
	documentation for making contributions to the project. The top five ranked
	software packages include information on how to contribute. 
	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs
	and changes. Issues should be regularly reviewed and closed. All but three
	(laboetie, lettuce, Sailfish) of the software packages that use Git have
	most of their issues closed.
	\item Use a version control system (GitHub, CVS). Four (ESPREsSo, Ludwig,
	LUMA, Palabos) of the top five ranked packages use a version control system.
	\item Source code needs to be well commented. Typically, more than 10
	percent of LBM package source code is comments, as presented in
	Table~\ref{gitrepodata}. Four of the top five overall ranked packages
	(Ludwig, ESPResSo, Palabos, OpenLB) have about 20 percent of their source
	code as comments.
	\item Modularize the source code, separate components, hide information
	behind well defined interfaces. This is suggested in SCS literature, and in
	developer comments.
	\item Eliminate code duplication. This is suggested as good software
	development practice. Duplicate code is difficult to maintain.
	\item If possible, consider using popular user-friendly software languages
	like Python. Especially consider this for parts of the source code that is
	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
	to address several software qualities, including maintainability.
	\item Consider the recommended points for addressing traceability.
\end{itemize}

@@

\begin{itemize}
	\item Modularize the source code, separate components, hide information
	behind well defined interfaces. This is suggested in SCS literature, and in
	developer comments.
	\item If possible, consider using popular user-friendly software languages
	like Python. Especially consider this for parts of the source code that is
	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
	to address several software qualities, including modifiability.
	\item Consider future source code modifiability as early as the design stage
	of development. This is suggested as good software development practice.
	\item Consider flexibility of data structures and data storage in the design
	stage. The package pyLBM redeveloped data structures to ease future changes.
\end{itemize}

@@

\begin{itemize}
	\item Modularize the source code, separate components, hide information
	behind well defined interfaces. This is suggested in SCS literature, and in
	developer comments.
	\item Document module interfaces in design and developer documentation. This
	is suggested as good software development practice.
	\item Provide API documentation, if applicable. Only one (ESPResSo) of the
	top five ranked packages provided API documentation.
\end{itemize}

@@

\begin{itemize}
	\item Update all relevant documentation when a change to the software is
	made. This is suggested as good software development practice.
	\item Use automatic document generation tools (Doxygen, Drasil, Sphinx) to
	limit the time spent on updating documentation. This was done by all of the
	top five ranked packages.
	\item Provide a developer's guide to help orient developers to the
	artifacts, source code, and system architecture so that they can better
	document changes. Top ranked package OpenLB has a developer guide. 
\end{itemize}

@@

\subsection{How does software designated as high quality by this methodology
compare with top-rated software by the community?} \label{comparison}

>>>>

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with Harvard style bibliographic references

@@

Seventeen of the software packages include a requirements specification artifact
or explicitly reference domain theory, often only the latter. Software packages
that distribute requirements specification information, such as DL\_MESO (LBE),
generally keep it brief and include it within other documentation. This artifact
is often found within a user manual, on a web-page, or is mentioned in related
publications. In the latter case the user may need to spend significant time to
find this information.  \wss{More on requirements as a pain point is given in
Weiss, and other papers cited by Carver (2021).  Isn't generally requirements in
the SE sense of following a template or similar.}

@@

\section{Comparison to Community Ranking} \label{repmetrics}

Table~\ref{repometrics} presents our LBM software package rankings along with
the repository ranking metrics of each software package. Nine packages do not
use GitHub, so they do not have a measure of repository stars. Looking at the
repository stars of the other 15 packages, we can observe a slight pattern where
packages that have been highly ranked by our assessment do have more stars than
lower ranked packages. The best ranked package (ESPResSo) has the second most
number of stars. The ninth ranked package (Sailfish) has the most number of
stars. Packages designated as lower quality often do not use GitHub or GitLab,
or have few stars. HemeLB only has 12 stars, but its location recently moved
from a different GitHub repository. Our assessment of this package might not be
accurate. The number of stars does not necessarily represent the perceptions of
the community, but for lack of an alternative measure we will use it this way.
Besides missing data, another threat to the validity of this comparison is the
varying ages of the repositories. Older packages have been able to accumulate
stars and watches for longer than newer packages. The true quality of new
packages may not be reflected in their stars and watches.

\begin{table}
\begin{center}
	\begin{tabular}{ p{3.5cm}p{2.5cm}p{2cm}p{2cm}p{2cm} }
		\toprule
		Name & Our Ranking & Repository Stars & Repository Star Rank & Repository Watches \\
		\midrule
		ESPResSo & 1 & 145 & 2 & 19\\
		Ludwig & 2 & 27 & 8 & 6\\
		Palabos & 3 & 34 & 6 & GitLab\\
		OpenLB & 4 & No Git & No Git & No Git\\
		LUMA & 5 & 33 & 7 & 12\\
		pyLBM & 6 & 95 & 3 & 10\\
		DL\_MESO (LBE) & 7 & No Git & No Git & No Git \\
		Musubi & 8 & No Git & No Git & No Git \\
		Sailfish & 9 & 186 & 1 & 41\\
		waLBerla & 10 & 20 & 9 & GitLab\\
		laboetie & 11 & 4 & 13 & 5\\
		TCLB & 12 & 95 & 3 & 16\\
		MechSys & 13 & No Git & No Git & No Git\\
		lettuce & 14 & 48 & 4 & 5\\
		ESPResSo++ & 15 & 35 & 5 & 12\\
		MP-LABS & 16 & 12 & 11 & 2\\			
		SunlightLB & 17 & No Git & No Git & No Git\\
		LB3D & 18 & No Git & No Git & No Git\\			
		LIMBES & 19 & No Git & No Git & No Git\\
		LB2D-Prime & 20 & No Git & No Git & No Git\\		
		HemeLB & 21 & 12 & 11 & 12\\
		lbmpy & 22 &  11 & 12 & 2\\	
		LB3D-Prime & 23 & No Git & No Git & No Git\\	
		LatBo.jl & 24 & 17 & 10 & 8\\			
		\bottomrule
	\end{tabular}
	\caption{Repository Ranking Metrics} \label{repometrics}
\end{center}
\end{table}

Table~\ref{rankingcomparison} compares the top 10 ranked LBM software packages
with a LBM package ranking made by a domain expert. Five (Palabos, OpenLB, LUMA,
Sailfish, waLBerla) of the top 10 ranked packages in this assessment are also
found in the domain expert's top 10 ranking. Interestingly, the first three of
these five packages are listed in the same order on both lists. Looking at the
remaining packages, the top two ranked packages in this state of the practice
assessment (ESPResSo, Ludwig) are not on the domain expert's list. Perhaps the
intended applications of these packages, complex fluids for Ludwig and soft
matter research for ESPResSo, do not align with the research interests of the
domain expert. This reasoning may also apply to eighth ranked Musubi. Moving
down the list, pyLBM and DL\_MESO(LBE) did not make the domain expert's top 10
list but were mentioned by the domain expert. Looking at the remainder of the
domain expert's list, ASL is a general purpose tool for solving partial
differential equations and may have fallen out of scope of the authoritative
lists that were used to identify the initial LBM software list. The domain
expert's fifth ranked package (ch4-project) was on the initial software list of
this assessment, but was removed due to a lack of documentation. Open FSI is a
new project that uses Palabos. It was made public the year before data was
collected for this state of the practice exercise. It may not have been on
authoritative lists due to its young age. Similarly, LIFE was also made public
recently. Finally, the domain expert's 10th ranked package (HemeLB) was ranked
21st in this state of the practice assessment.

@@

@@

This section answers the research questions listed in Section~\ref{secIntro}
using the quantitative data presented in Section~\ref{AHPresults}, qualitative
data presented in Section~\ref{interviewresults}, additional data from software
package repositories and artifacts, and domain expert feedback. Each subsection
answers one of the research questions. Software package artifacts are examined
in Section~\ref{artifacts}, tools are listed in Section~\ref{tools}, development
principles, processes, and methodologies are discussed in Section
~\ref{prinprocmeth}, development pain points are explored in Section
~\ref{painpoints}, and summary of current (Section~\ref{currentPractices}), best
(Section~\ref{bestPractices}) and future practices
(Section~\ref{futurePractices}). Finally, threats to the validity of this
assessment are noted in Section~\ref{threats}.

@@

\begin{table}
\begin{center}
\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
\toprule
\textbf{Common} & \textbf{Less Common} & \textbf{Rare} \\
\midrule
Authors / Developers List & Change Log / Release Notes & API Documentation\\
Bug Tracker & Design Documentation & Developer / Contributor Manual\\
Dependency Notes & Functional Spec. / Notes & FAQ / Forum\\
Installation Guide / Instructions & Performance Information & Verification and Validation Plan\\
Requirements Spec. / Theory Notes & Test Plan / Script / Cases & Video Guide (including YouTube)\\
List of Related Publications & User Manual/Guide & \\
Makefile / Build File & Version Control & \\
README File & & \\
License & & \\
Tutorial & & \\
\bottomrule
\end{tabular}
\caption{Artifacts Present in Surveyed Packages, Classified by Frequency}
\label{artifactspresent}
\end{center}
\end{table}

@@

The top four AHP ranked packages have most of the less common artifacts. At the
time of data collection, only one of the four packages (Palabos) did not have a
user manual or guide, but there was a broken link on the package website
indicating that such an artifact might exist. This broken link was later fixed,
but this is not reflected in our data because it was not present at the time of
data collection. Despite the broken link, Palabos does have a detailed and
informative website. Another one (Ludwig) of the top four packages does not
appear to have publicly visible design documentation. A third package (OpenLB)
from the list does not appear to use a version control system. It is possible
that such a system is used by the package, since its website notes package
version numbers, but the artifacts do not explicitly state the use of such a
system.

@@

\wss{Weiss mentions pain points related to design documentation.  Carver (2021) 
also mentions design documentation in the literature review.}

@@

A developer of pyLBM noted that their small development team has a lack of time
to implement new features. Small development teams are common for LBM software
packages. Team members are almost always part of the same institute or already
know each other from other projects. External contributions are rare despite
many of the projects accepting them. \wss{Other projects cite the same point
point - Wiese paper}  Aside from on-boarding new developers, time constraints
could be mitigated by increasing developer efficiency, which could be addressed
in several ways, including by improving the quality of documentation, or
incorporating automatic code generation. \wss{Potential ways to address the pain
points might go into a separate section?  Maybe propose the strategy, like code
generation, and then list the pain points that are addressed?}

@@

A lack of software development experience was noted by the developer of TCLB.
Many of the team members on their project are domain experts and there can be a
steep learning curve before these members contribute good quality source code.
This problem has been mitigated by re-writing the code to more easily enable
future contributions.  On a related point, improving the software engineering
education or experience of developers is also an idea that was brought up by
several developers.  \wss{The Wiese paper and others talk about the background
of developers.}

@@

The TCLB developer noted a lack of incentive and funding in academia for
developing widely used scientific software. The importance of funding for
scientific software has been discussed in \cite{gewaltig2012quality}, which
notes that ``software tools are developed and maintained only for as long as
there is explicit or implicit funding''. The developer further commented that
there are no journals that publish such scientific software source code.
However, there are ways to get such source code cited. Work has been done to
address this in \citep{smith2016software}, which presents a set of software
citation principles and discusses ``how they could be used to implement software
citation in the scholarly community'' \citep{katz2019software}.  \wss{other
papers also mention this pain point.  Solutions proposed by Katz, journal of
open source etc.}

@@

Scientific software is often developed in-house by the very researchers that
temporarily use it in their own research. Empirical studies of such
``professional end-user development'' of SCS is noted in \citep{segal2007end}.
This kind of software has a defined user and purpose, and often does not meet
the standards that would be required by external users. It has been categorized
as a ``private tool'' by \citep{gewaltig2012quality}, which notes that despite
often being made freely available, ``it is not always clear that it is
sufficiently mature in terms of domain coverage, validity, documentation or
usability, to be useful to other researchers''. It is less common for such ad
hoc scientific software to become ``user-ready software'', which ``is not only
research-ready, but should have most of the attributes commonly expected of
commercial software products including broadness of scope, robustness,
demonstrable correctness and adequate documentation''
\citep{gewaltig2012quality}. As software becomes user-ready, it can become
commercialized and closed-source.  \wss{remove this paragraph?}

\wss{Remove this section?  The external support organizations will likely come
up as the mitigation for some of the other pain points.}

@@

Setting up parallelization was also noted as a technical pain point by one of
the developers, and the introduction of continuous integration by another.
Software development knowledge, and automatic code generation, could mitigate
such pain points. As already noted, many of the developers are domain experts
and not professional software developers. The developer of TCLB noted that
eliminating equivalent statements using macros had helped improve the quality of
their source code, specifically helping with reusing code to run on both the CPU
and GPU. \wss{The point about macros isn't really related to parallelization or
continuous integration.}

@@

Difficulties with ensuring correctness were also noted by several developers.
They indicated that tests are run on all new source code additions, testing both
individual modules and the system to verify correctness. These tests compare the
package output to known correct output using test cases. The developer of TCLB
commented that the amount of testing data that is needed for some cases is a
problem as free testing services do not offer capabilities to store and process
such large amounts of data, and in-house testing solutions needed to be created
to address this limitation. The solution for this has been to limit the size of
the testing problems, and to run tests in small batches with few iterations.
\wss{The Weiss paper alludes this this as well.  I think the Kelly and Sanders
paper likely talks about testing problems and correctness.}

@@

A few obstacles related to users were found. Several developers noted that users
sometimes try to use incorrect LBM method combinations to solve their problems.
Furthermore, some users think that the packages will work out of the box to
solve their cases, while the packages require both a good understanding of CFD
and an understanding of the requirements for formulating problems in the
individual packages, which can be a significant endeavor. These software
packages are not like commercial software packages. They are generally set up to
solve specific research problems, and are often primarily used by their
developers. While they are modifiable to solve similar problems, these
modifications are not trivial. Better documentation, with attention on
traceability, and automatic code generation are suggested when designing
software for change, and would help with these modifications. \wss{Should decide
if want to introduce ideas for improving within the presentation of the pain
point, or separately?}  So far this problem of on-boarding new users has been
addressed by updating the documentation to better inform users of the underlying
LBM theory and package requirements. Similar issues with LBM parameters were
noted by another developer. Updating the user interface to better explain
theoretical principles, as well as test user input for compatibility, was the
implemented solution. As noted above, sometimes frequently asked questions on
the underlying theory and on how to use the software are answered in the
documentation.  The interviewed developer of ESPResSo commented that parts of
their package's source code had been refactored to Python to help address
usability issues. Python was perceived as a much more usable language, and it
would be easy for future users and developers to learn and understand the source
code.

@@

A few potential future obstacles were noted. The developer of ESPResSo noted
that their source code had been written with a specific application in mind and
that due to this there was too much coupling between components of the source
code. This results in technical debt, having an impact on future modifiability
and reusability when trying to extend the software, and the code would need to
be refactored.  As noted above, difficulties with ensuring future correctness
could also arise. As new methods and functionality is added into the software,
new test cases and test data will need to be developed.  \wss{Weiss mentions
mainainability related issues under pain points.}

@@

Interviewees commented that documentation is important and that its quality
could be improved. As already noted, there is often no time or funding for
maintaining quality documentation for software that is rarely used outside of
the development team. Furthermore, the documentation generally only provides a
shallow overview of the underlying CFD theory. Users would be well advised to
already be familiar with these topics, or they should spend significant time
referencing theory resources. The documentation instead generally focuses on
explaining how to use the software. It is of course not feasible for package
documentation to address the underlying physics topics in detail, so it is
advised that the package documentation links to resources that better explain
the underlying theory. Sometimes, frequently asked questions about the
underlying theory are answered in the documentation. OpenLB has an artifact for
such questions.  \wss{Weise mentions documentation.}

@@

\subsection{Current Practices} \label{currentPractices}

The following points regarding software quality should be considered when
developing LBM software packages. These points are based on developer
interviews, SCS literature, what was found to have worked for packages that were
designated as high quality in this assessment.  Our recommendations are not
lists of what should have been done in the past, or what should be done now;
they are just suggestions for consideration in the future.  \wss{Rework this
section to highlight what the top projects are doing.  Point to checklists for
other advice?}

@@

\begin{itemize}
	\item Include OS compatibility, including specific OS versions. This was
	done by several top ranked packages, including ESPResSo and LUMA.
	\item Provide complete installation instructions. This was done by all of
	the top five ranked packages.
	\item Installation instructions should be written as if the user does not
	have any dependencies installed and is installing on a clean OS. This was
	done by several top ranked packages, including ESPResSo, OpenLB, and
	Palabos. This can be tested on a clean environment using a virtual machine.
	\item List all dependencies in the installation instructions, dependency
	versions, and how to install them. This was done by several top ranked
	packages, including ESPResSo, LUMA, OpenLB, and Palabos.
	\item If possible, automate the installation of dependencies. Use tools such
	as Make. This was done by all of the top five ranked packages.
	\item Automate the installation process as much as possible. Use tools such
	as Make. This was done by all of the top five ranked packages.
	\item The installation instruction should only be in one location. This was
	done by several top ranked packages, including ESPResSo, OpenLB, and
	Palabos.
	\item Include descriptive error messages for errors encountered during
	installation. This was done by several top ranked packages, including Ludwig
	and LUMA.
	\item Provide a way to validate the installation. This can be done using a
	custom script, or a test case that specifies expected output. This was done
	by several top ranked packages, including OpenLB, Ludwig, LUMA, and Palabos.
	\item Provide instructions for uninstalling the software. These were
	included with pyLBM.
\end{itemize}

@@

\begin{itemize}
	\item Use a requirements specification document. This is suggested in SCS
	literature, and several of the top ranked packages had such a document or
	reference to theory manuals. A potential template is presented in
	\citep{smith2005new}. 
	\item Make public (on package website or GitHub) the requirements
	specification document, or explicitly reference the domain theory that the
	software is designed from.  This was done by several top ranked packages,
	including ESPResSo, Ludwig, LUMA, and OpenLB.
	\item Ensure the above information is easy to find. Consider adding it to
	the user manual. The information is easy to find in most top ranked
	packages.
	\item Development teams should include both domain experts and experienced
	software developers. This suggestion is based on developer comments.
	\item Use and make public (on package website or GitHub) detailed
	documentation. Consider using automatic document generation tools like
	Doxygen, Drasil, or Sphinx. This was done by all of the top five ranked
	packages.
	\item Provide detailed tutorials that include expected output, like the
	\href{https://www.walberla.net/doxygen/index.html}{waLBerla tutorials}.
	\item Use unit tests during development and make them public (on package
	website or GitHub). This was available for Ludwig.
	\item Modularize the source code, separate components, hide information
	behind well defined interfaces. This is suggested in SCS literature, and in
	developer comments.  \wss{one of Wilson's best practices.}
	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and
	processes during development. This was done by top ranked packages ESPResSo
	and Ludwig.  \wss{Common in LBM - more common than in other domains}
\end{itemize}

@@

\begin{itemize}
	\item Include descriptive error messages where appropriate. This was done by
	most packages that encountered a fault.
	\item In case automatic installation of dependencies fails, the system
	should indicate to the user what dependencies need to be installed manually.
	This was done by many packages, including ESPResSo++, Ludwig, LUMA, pyLBM,
	TCLB.
	\item The packages should include detailed tutorials, including
	dependencies, expected output, and any additional supplementary
	documentation that may be required. This was done by many packages,
	including waLBerla, Palabos, MechSys, LUMA, and pyLBM.  \wss{Comment from
	ASCR attendee that success with open source requires good tutorials and
	documentation - is this published anywhere?}
\end{itemize}

@@

\begin{itemize}
	\item Keep artifacts updated. This is suggested as good software maintenance
	practice.
	\item Ideally include most of the common and less common artifacts listed in
	Section~\ref{artifacts}.
	\item Include version numbers and release notes for all major source code
	and artifact releases. The top five ranked software packages include release
	notes.
	\item Have a defined process for accepting contributions, and make public
	documentation for making contributions to the project. The top five ranked
	software packages include information on how to contribute. 
	\item Use an issue tracker (Git, email, SourceForge, other) to manage bugs
	and changes. Issues should be regularly reviewed and closed. All but three
	(laboetie, lettuce, Sailfish) of the software packages that use Git have
	most of their issues closed.
	\item Use a version control system (GitHub, CVS). Four (ESPREsSo, Ludwig,
	LUMA, Palabos) of the top five ranked packages use a version control system.
	\item Source code needs to be well commented. Typically, more than 10
	percent of LBM package source code is comments, as presented in
	Table~\ref{gitrepodata}. Four of the top five overall ranked packages
	(Ludwig, ESPResSo, Palabos, OpenLB) have about 20 percent of their source
	code as comments.
	\item Modularize the source code, separate components, hide information
	behind well defined interfaces. This is suggested in SCS literature, and in
	developer comments.
	\item Eliminate code duplication. This is suggested as good software
	development practice. Duplicate code is difficult to maintain. \wss{No
	support for this one or citation, either find or remove it.}
	\item If possible, consider using popular user-friendly software languages
	like Python. Especially consider this for parts of the source code that is
	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
	to address several software qualities, including maintainability.
	\item Consider the recommended points for addressing traceability.
\end{itemize}

@@

\begin{itemize}
	\item Modularize the source code, separate components, hide information
	behind well defined interfaces. This is suggested in SCS literature, and in
	developer comments.  \wss{add citation - Wilson mentions - this points seems
	repetitive}
	\item Consider future source code modifiability as early as the design stage
	of development. This is suggested as good software development practice.
	\wss{citation, support, or remove.}
	\item Consider flexibility of data structures and data storage in the design
	stage. The package pyLBM redeveloped data structures to ease future changes.
\end{itemize}

@@

\begin{itemize}
	\item Modularize the source code, separate components, hide information
	behind well defined interfaces. This is suggested in SCS literature, and in
	developer comments.  \wss{repeated from elsewhere - there should be an
	easier way to do this - give the advice and then link it to the qualities,
	instead of this way?}
	\item Document module interfaces in design and developer documentation. This
	is suggested as good software development practice. \wss{citation or support}
	\item Provide API documentation, if applicable. Only one (ESPResSo) of the
	top five ranked packages provided API documentation.
\end{itemize}

@@

\begin{itemize}
	\item Update all relevant documentation when a change to the software is
	made. This is suggested as good software development practice.
	\wss{support, cite or remove}
	\item Use automatic document generation tools (Doxygen, Drasil, Sphinx) to
	limit the time spent on updating documentation. This was done by all of the
	top five ranked packages. \wss{repetitive}
	\item Provide a developer's guide to help orient developers to the
	artifacts, source code, and system architecture so that they can better
	document changes. Top ranked package OpenLB has a developer guide. \wss{How
	is this related to traceability?}
\end{itemize}

@@

\subsection{Best Practices} \label{bestPractices}

practices by the best measured.

practices mentioned in literature, but not practiced (if any)

\subsection{Future Practices} \label{futurePractices}
